Moisture effects on leaf litter are species-specific and result in non-additive effects flammability in mixed conifer forest

### 2019-05-13 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Get extrafont working, edits

@@

@@

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Affiliation 1; e-mail@e-mail.com\\
$^{2}$ \quad Texas Tech UnNiversity; dylan.schwilk@ttu.edu}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{ Moisture content is a strong determinant of forest fuel
  flammability. But less is known about the mechanisms by which litter traits
  influence litter moisture dynamics, particularly the processes of moisture
  absorption and desorption. Characterizing how species’ leaf litter retain and
  release moisture through time is critical to the assessment of fuel’s
  availability to burn. Natural forests are often comprised of mixed stands,
  leading to mixtures of leaf litter on the surface fuelling the fires. Two
  studies have already established the existence of non-additive effects in the
  flammability of dry litter, but we lack a mechanistic understanding of
  moisture synergies in mixed litter and of moisture effects on the
  flammability of mixed litter. The litter from eight species of the
  mixed-conifer forest was saturated and allowed to dry to determine moisture
  absorption and desorption. Burn trials were performed in moist litter beds of
  single species and of mixtures, to establish flammability response to the
  dry-down process. Moisture absorption is determined by traits pertaining to
  leaf geometry while moisture desorption is influenced by emergent properties
  of the litterbed. Species that produce more aerated litter beds (such as Q.
  kelloggii and the pines) have faster rates of litter desiccation. My results
  show a relationship between time since wetting and time to ignition and
  spread rate, with significant interaction with genus. This indicates a
  genus-specific behavior of flammability and time since wetting. I found
  non-additive effects in litter mixtures for both moisture content and
  flammability. Mixtures containing Q. kelloggii and P. jeffreyi, species that
  dry faster, produced more flammable behavior (lower time to ignition, faster
  spread rate, taller flames). Fire behavior of mixtures depended on moisture
  content, although mixtures without Q. kelloggii had always higher moisture
  content than predicted by the average fire behavior of the species in the
  mixture. At lower moisture contents (<75\%) the mixtures exhibit positive
  non-additive behavior and at higher moisture contents (> 200\%) they exhibit
  negative non-additive behavior.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setcounter{section}{-1} %% Remove this when starting to work on the template.
\section{How to Use this Template}
The template details the sections that can be used in a manuscript. Note that the order and names of article sections may differ from the requirements of the journal (e.g., the positioning of the Materials and Methods section). Please check the instructions for authors page of the journal to verify the correct order and names. For any questions, please contact the editorial office of the journal or support@mdpi.com. For LaTeX related questions please contact latex@mdpi.com.
%The order of the section titles is: Introduction, Materials and Methods, Results, Discussion, Conclusions for these journals: aerospace,algorithms,antibodies,antioxidants,atmosphere,axioms,biomedicines,carbon,crystals,designs,diagnostics,environments,fermentation,fluids,forests,fractalfract,informatics,information,inventions,jfmk,jrfm,lubricants,neonatalscreening,neuroglia,particles,pharmaceutics,polymers,processes,technologies,viruses,vision

@@

Fuel moisture content is an important, and often driving, determinant of surface fire behavior. Environmental conditions such as temperature, relative humidity, and wind, drive the moisture content of the fuel available to burn [CITE]. In particular, fine dead fuels respond very readily to environmental changes [CITE]. In mixed-conifer forest of the western USA, where the fire regime is characterized by surface fires fuelled by plant litter, leaf traits can influence landscape scale fire behavior (Schwilk \& Caprio 2011). Litter packing (measured as litterbed density or packing ratio) controls the flammability of dry litter fuels in seasonally dry climates (higher litter density reduces flammability) and leaf litter packing is influenced by leaf size and shape (Fonda 1998, Kane et al. 2008, Schwilk \& Caprio 2011, Kreye et al. 2013). Fine fuel moisture content strongly reduces ignition probability and flame spread rate (Gisborne 1936, Fons 1946, Anderson \& Rothermel 1965, Xanthopoulos \& Wakimoto 1993, Dimitrakopoulos \& Papaioannou 2001, Weise et al. 2005, Pellizzaro et al. 2007). However, the mechanism by which both monospecific and mixed litterbeds absorb and desorb moisture is poorly known, and could be characterized by leaf traits, litter traits, or both.

Fuel moisture influences flammability and fire behavior through effects on ignition, spread rate, flame height, fuel consumption, and heat release, due to the high specific heat content of water (Rothermel 1972). Specifically, high moisture contents decrease the likelihood of ignition thereby reducing fire spread (Rothermel 1983, Nelson 2001). Moisture acts as a heat-sink resulting in less energy available for propagation, and the resulting water vapor will both cool the flaming front and dilute flammable gases (Albini 1976, Shafizadeh et al. 1977). Variation in moisture of living fuels has strong effects on flammability, but moisture of dead excised litter fuels can also vary. Moisture of litter fuels varies with climate, with seasons and with short term weather within a season.  Therefore, the rate of drying and flammability at any particular moisture content are key parameters that influence the fuel flammability throughout the fire season.  If species differ in the rate of drying following a precipitation event, this may strongly influence relatively flammability across species at any particular time post-wetting.  Such differential moisture content across time, may contribute to the often high spatial heterogeneity of surface fires (Pickett et al. 1999, Keeley et al. 2009). Fuel moisture dynamics of leaf litter are probably dependent upon forest structure (shading and rain throughfall effects), litter physical structure, and chemical composition of the fuelbeds (Swift et a. 1979, Hille \& Stephens 2005, Nelson \& Hiers 2008, Matthews 2014).

Fuel moisture content through time is controlled by two main processes, absorptive capacity and rate of moisture loss (Kreye et al. 2013). Absorptive capacity represents the potential maximum fuel moisture and is evaluated by the moisture content of litter when saturated. After saturation, the differences in moisture content indicate different abilities to retain moisture. The ability of litter to absorb and desorb water should be dependent on leaf and/or litter geometry, however, which traits control these processes are largely unknown. I propose two different sets of traits responsible for variation in the two processes mentioned: leaf particle geometry and litter bulk density. Litter characteristics such as surface-area-to-volume ratio, litter bulk density (Gibos 2010), mass density, particle size and shape, and chemical traits (Nelson 2001) influence moisture diffusion through litter. Litter packing has two pathways of effect on fire behavior: directly by its effect on oxygen availability for combustion, and indirectly by its effect on dry-down rate, and both these pathways are affected by leaf litter traits (such as leaf size, litter bulk density). Fuelbeds composed of larger leaves with lower moisture retainability and decomposition rates favor drying, thus increasing the probability of ignition and fire spread rate. 
Species traits may interact such that the flammability of multi-species mixtures is not the average of the constituent species (van Altena et al. 2012, de Magalhães \& Schwilk 2012, Blauw et al. 2015).  Such non-additive effects have been shown to be synergistic in dry litter fuels where the flammability of a mixture is driven by the most flammable species (de Magalhães \& Schwilk 2012, van Altena et al. 2012). In most forests, leaf litter will be a mixture representative of the species in the stand. It is possible that similar non-additive effects may influence litter moisture content, especially if different leaf traits influence adsorptive capacity and drying rate. Recent work by Blauw et al. (2015) suggests that moisture content can contribute to non-additive effects on heathland flammability, but the mechanism of this effect is not known. I investigated non-additive effects on both moisture content and the flammability of litter mixtures varying in time since wetting and in moisture content. I expect that litter mixtures will differ in their bulk density and thus, will dry at different rates.

I investigated species trait effects on litter moisture dynamics and consequent litter flammability across eight tree species in a temperate fire-prone forest in California. I investigated the relationship between traits such as leaf thinness and litter bulk density with the processes of moisture absorption and desorption. I determined the rate of moisture loss, the moisture of extinction (the threshold representing the moisture value above which fire will not ignite or spread) and the influence of moisture on the flammability of eight species from a mixed-conifer forest in California. I performed burn trials on leaf litter at various moisture levels and determined how time since wetting influenced time to ignition and flame spread rate. The burn trials were also performed with mixtures to evaluate if species contribute equally to litter moisture dynamics or if there are non-additive effects. I studied whether the non-additive effects on flammability reported in previous studies (de Magalhães \& Schwilk 2012, van Altena et al. 2012) occur under ecologically-relevant moisture levels and whether fuel moisture effects strengthen or mask such non-additivity.

@@

Field sites for this study were located in Sequoia and Kings Canyon National Parks, California, USA (36 36N, 118 42 W) between 1600 and 2400 m elevation in mixed-conifer forest. I chose eight tree species that are representative of this type of forest: \emph{Pinus jeffreyi} Grev. \& Balf., \emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws., \emph{Abies concolor} (Gord. \& Glend.) Lindl. ex Hildebr., \emph{Abies magnifica} A. Murr., \emph{Calocedrus decurrens} (Torr.) Florin, \emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J. Buchholz. In this study, I consider litter the top layer of leaves and small twigs less than 0.625 cm diameter (= 1-hour fuel, Fosberg 1970) that have fallen in the previous year (mostly undecomposed). I collected the leaf litter in the summer (mid-June to mid-July) in 2011 and in 2012 from 21 separate sites across the study area. Effort was made to collect from different populations (minimum of four, each population separated from the others by 9 to 32 km) for each species to control for plastic and ecotypic effects. An exception to this involved the litter from A. magnifica which is restricted to the higher elevations and could only be collected from one area, due to logistical constraints. Within that area, \emph{A. magnifica} litter was collected from over 10 different individuals. For the remaining species, the collection involved 2-4 individual trees at each site, separated by at least 10 m. The litter was collect about 2 m away from the trunk to obtain a more uniform sample, because bark and heavier branches tend to fall closer to the tree trunk.

@@

I produced dry-down curves to determine maximum water retainability (moisture absorption) and the rate of moisture content loss (moisture desorption) for each species and for four distinct mixtures. These curves allowed me to determine the rate at which a species loses moisture (approximated as exponential decay), and determine maximum water content following saturation and gravity draining.  Samples from the eight species (six replicates per species) and four mixtures (5 replicates) were placed in baskets constructed of aluminium screen. I used 450 g of litter (dry weight) per trial, and controlled for depth, by changing the size of the basket. Weight was measured by using a balance sensitive to 0.1 g (model XS16001L, Mettler Toledo, Columbus, OH). Samples were immersed in clear plastic storage bins filled with water (Figure 2.1) for 24 hours to fully saturate them. After saturation, the baskets were removed from the water, allowed to drain for 3 minutes, and samples weighed to measure maximum moisture content. The samples were then allowed to dry in a partially controlled environment kept at 21 C temperature and 30\% relative humidity, assessed via an iButton sensor (Maxim Integrated) placed over the samples.  During the dry-down process three subsamples were taken at 24 hour intervals and weighed. These subsamples were then oven dried for 24 hours at 100 C and re-weighed to assess moisture content. This was repeated until the final sample from a basket was less than10\% moisture.

 The dry-down curves indicated that the eight species varied in their moisture behavior (Figure 2). To determine a subset of species to use in the mixture trials, I performed analysis of contrast using lsmeans (package: lsmeans; Lenth 2016) to determine similarity between the species, which fell into four groups. I then selected one species from each group to include in the mixtures resulting in four distinct 3-species mixtures.

Leaf trait data
We investigated the influence of leaf thinness and litter bulk density on the moisture dynamics of the species in this study. Leaf thinness data (leaf length over thickness) was obtained for subsamples of the litter collected for this project.  Bulk density of the litter was obtained for each basket during the dry-down experiment by dividing the dry mass (450 g) by the estimated litter volume (basket width x length x litter depth).

@@

I performed flammability tests across multiple moisture levels on monospecific litter beds and on mixed litter beds replicated 5 times. I measured flammability using a 150 cm long burn table, 15 cm wide and 15 cm tall, in which leaf litter was placed and the table gently shaken to allow settling. In my experience, this design allows fire behaviour to reach constant flame spread rates (de Magalhães and Schwilk 2012).  Two tall metal rulers equally spaced along the apparatus allow visual determination of maximum flame height and nine type K thermocouples attached to data loggers measured temperature every second at three distances from the ignition point and at three different depths below the litter surface. Litter was ignited with a single ignition source (propane torch; maximum adiabatic flame temperature of 1899 C, http://www.benzomatic.com/products/fuel.aspx) at one end of the apparatus and allowed to burn until extinction. If a species at a specific moisture level failed to ignite after three attempts, no further attempts were made.  A timer was used to record time to ignition and duration of flaming combustion. The flammability trials followed the methodology detailed on de Magalhães and Schwilk (2012).

During flammability trials, I measured time to ignition (s), duration of flaming combustion (s), spread rate (mm/s), calculated by dividing the length of the burned surface by the time it takes the flaming front to reach the end of the apparatus, maximal flame height (mm), and mass loss (\%), calculated as the difference between fuel mass before and after flame extinction (Anderson 1970, Cornelissen et al. 2003, Ormeño et al. 2009). I also measured duration above 100 C, or the length of time temperature exceeded 100 C, and temperature integration, which integrates duration of combustion and temperature above 100 C and serves as a proxy for heat release (de Magalhães and Schwilk 2012, Cornwell et al. 2015). By performing the flammability tests at several moisture levels I were able to determine moisture of extinction (a threshold representing the moisture value above which litter will not ignite or spread) for all species in this study, using a logistic regression.

Burn trials occurred in a cement structure used to simulate house fires at the Fire Department of the City of Lubbock.  Though not completely air tight, this structure minimizes wind and helps regulate temperature and relative humidity. Temperature and relative humidity were measured before every trial using a Kestrel 3000 (Nielsen-Kellerman, Boothwyn, PA). To minimize temperature and relative humidity variation, I conducted trials between 10 am and 3 pm.

@@

Dry-down curves (Figure 2.2) were created based on all data points collected, fitted with an exponential curve to each species:

%m = mmax e-λt								Eq. 1

Were m is moisture content at time t, and $\lambda$ is the exponential decay constant. I used a liner mixed-effects model taking the natural logarithm of moisture content, with time since wetting and species as fixed effects and nested within replicate. I extracted the coefficients of the fitted model to describe maximum water retainability (mmax; g of water per g of dry mass) and desiccation index ($\lambda$) for each species. I used post-hoc Tukey tests (package: lsmeans; Lenth 2016) to discriminate species’ behavior in terms of maximum water retainability and desiccation index. I analysed the influence of leaf size and litter bulk density on moisture dynamics (maximum water retainability and desiccation index) using linear nested models and used anova to determine which traits influenced the two processes. The effect of time since wetting on flammability was modelled using linear regression with the interaction between time since wetting and genus specified and with average vapor pressure deficit in the burning room during that trial as predictor. Differences across species were tested with post-hoc Tukey tests. A binomial logistic regression was fitted to the probability of ignition to obtain moisture of extinction.
I investigated non-additive effects on moisture and on flammability by testing if moisture content and flammability of mixtures were predicted by the average of the constituent species. I created the predicted values using the models that determine the relationship between moisture content/flammability and time since wetting for individual species, and then averaging the results obtained for the species in a mixture. This procedure was done for both moisture content and flammability parameter values of the constituent species. The expectation from the null model is that the difference between observed and predicted values of flammability will be zero. The null model is created by averaging the moisture content and the flammability of the individual species comprising the mixture for each time since wetting measured. I tested whether there were significant departures from the null (zero) with a mixed effects linear model (package: lme4; Bates et al. 2015). All analyses were conducted in R v.3.3.1 (R Core Team 2016).

@@

There are marked differences in the extent of moisture species retain, seen as the maximum moisture content at time 0 (y-axis intercept, Fig. 2.2). Q. kelloggii, the only broadleaf, is the species that retains more moisture. Tukey tests on the linear mixed-effects model discriminated the species into three different maximum water retention groups: 1) Q. kelloggii; 2) P. lambertiana and C. decurrens; and 3) A. magnifica, S. giganteum, A. concolor and P. jeffreyi. P. ponderosa could not be discriminated between the last two groups. There are also marked differences in the desiccation rate of the eight species and the Tukey tests discriminated species into three groups: 1) Pinus; 2) the Cupressaceae; and 3) Q. kelloggii in between. C. decurrens is among the species with the lowest moisture retainability and is the species with the slower desiccation. Species with lower litter bulk density, such as Q. kelloggii and the pines, have higher desiccation rates than all other species.

The regression between maximum water retention and leaf litter traits showed that maximum water retention was significantly affected by a measure of leaf size (particle length/thickness; p=0.0369) but not by a characteristic of the litter bed (bulk density), indicating that the ability to absorb water is particle-dependent and a feature of a leaf’s morphology and/or chemistry (Table 2.1). Desiccation rate was significantly affected by litter bulk density but not leaf thinness (p=0.8794), indicating that desiccation rate is controlled by the emergent properties of the fuelbed. This is explained by the fact that more aerated fuel beds will facilitate evaporation, due to increased diffusion of water molecules from the leaf surface.

@@

Time since wetting was modelled against time to ignition and spread rate using genus as a fixed effect, because in a previous study (de Magalhães and Schwilk 2012) I observed that species flammability grouped around genus. Both models showed significant effects

@@

@@

\begin{figure}[H]
\centering
\includegraphics[width=2 cm]{Definitions/logo-mdpi}
\caption{This is a figure, Schemes follow the same formatting. If there are multiple panels, they should be listed as: (\textbf{a}) Description of what is contained in the first panel. (\textbf{b}) Description of what is contained in the second panel. Figures should be placed in the main text near to the first time they are cited. A caption on a single line should be centered.}
\end{figure}

>>>>

%% reftex-default-bibliography: ("/home/schwilk/write/bib/schwilk.bib")

@@

%\bibliography{/home/schwilk/write/bib/schwilk.bib}

@@

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Affiliation 1; e-mail@e-mail.com\\
$^{2}$ \quad Texas Tech University; dylan.schwilk@ttu.edu}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{ Moisture content is a strong determinant of forest fuel
  flammability. But less is known about the mechanisms by which litter traits
  influence litter moisture dynamics, particularly the processes of moisture
  absorption and desorption. Characterizing how species’ leaf litter retain and
  release moisture through time is critical to the assessment of fuel’s
  availability to burn. Natural forests are often comprised of mixed stands,
  leading to mixtures of leaf litter on the surface fuelling the fires. Multiple
  studies have established the existence of non-additive effects in the
  flammability of dry litter, but we lack a mechanistic understanding of
  moisture synergies in mixed litter and of moisture effects on the
  flammability of mixed litter. The litter from eight species of a
  mixed-conifer forest was saturated and allowed to dry to determine moisture
  absorption and desorption. Burn trials were performed in moist litter beds of
  single species and of mixtures, to establish flammability response to the
  dry-down process. Moisture absorption is determined by traits pertaining to
  leaf geometry while moisture desorption is influenced by emergent properties
  of the litterbed. Species that produce more aerated litter beds (such as \emph{Q.
  kelloggii} and \emph{Pinus} species) have faster rates of litter desiccation. My results
  show a relationship between time since wetting and time to ignition and
  spread rate, with significant interaction with genus. This indicates a
  genus-specific behavior of flammability and time since wetting.
  We found non-additive effects in litter mixtures for both moisture content and
  flammability. Mixtures containing \emph{Q. kelloggii} and \emph{P. jeffreyi}, species that
  dry faster, produced more flammable behavior (lower time to ignition, faster
  spread rate, taller flames). Fire behavior of mixtures depended on moisture
  content, although mixtures without Q. kelloggii had always higher moisture
  content than predicted by the average fire behavior of the species in the
  mixture. At lower moisture contents (<75\%) the mixtures exhibit positive
  non-additive behavior and at higher moisture contents (> 200\%) they exhibit
  negative non-additive behavior.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \setcounter{section}{-1} %% Remove this when starting to work on the template.
% \section{How to Use this Template}
% The template details the sections that can be used in a manuscript. Note that the order and names of article sections may differ from the requirements of the journal (e.g., the positioning of the Materials and Methods section). Please check the instructions for authors page of the journal to verify the correct order and names. For any questions, please contact the editorial office of the journal or support@mdpi.com. For LaTeX related questions please contact latex@mdpi.com.
%The order of the section titles is: Introduction, Materials and Methods, Results, Discussion, Conclusions for these journals: aerospace,algorithms,antibodies,antioxidants,atmosphere,axioms,biomedicines,carbon,crystals,designs,diagnostics,environments,fermentation,fluids,forests,fractalfract,informatics,information,inventions,jfmk,jrfm,lubricants,neonatalscreening,neuroglia,particles,pharmaceutics,polymers,processes,technologies,viruses,vision

@@

Fuel moisture content is an important, and often driving, determinant of surface fire behavior. Environmental conditions such as temperature, relative humidity, and wind, drive the moisture content of the fuel available to burn [CITE]. In particular, fine dead fuels respond very readily to environmental changes [CITE]. In mixed-conifer forest of the western USA, where the fire regime is characterized by surface fires fuelled by plant litter, leaf traits can influence landscape scale fire behavior \cite{Schwilk+Caprio-2011}. Litter packing (measured as litterbed density or packing ratio) controls the flammability of dry litter fuels in seasonally dry climates (higher litter density reduces flammability) and leaf litter packing is influenced by leaf size and shape \cite{Fonda+Belanger+etal-1998, Kane+Varner+etal-2008, Schwilk+Caprio-2011, Kreye+Varner+etal-2013}. Fine fuel moisture content reduces ignition probability and flame spread rate \cite{Gisborne-1936, Fons-1946, Anderson+Rothermal-1965}. However, the mechanism by which both monospecific and mixed litterbeds absorb and desorb moisture is poorly known, and could be characterized by leaf traits, litter traits, or both.

Fuel moisture influences flammability and fire behavior through effects on ignition, spread rate, flame height, fuel consumption, and heat release, due to the high specific heat content of water \cite{Rothermel-1972}. Specifically, high moisture contents decrease the likelihood of ignition thereby reducing fire spread \cite{Nelson-2011}. Moisture acts as a heat-sink resulting in less energy available for propagation, and the resulting water vapor will both cool the flaming front and dilute flammable gases \cite{Albini-1976, Shafizadeh-1977}. Variation in moisture of living fuels has strong effects on flammability, but moisture of dead excised litter fuels can also vary. Moisture of litter fuels varies with climate, with seasons and with short term weather within a season. Therefore, the rate of drying and flammability at any particular moisture content are key parameters that influence the fuel flammability throughout the fire season \cite{Viney-1991}. If species differ in the rate of drying following a precipitation event, this may strongly influence relatively flammability across species at any particular time post-wetting.  Such differential moisture content across time, may contribute to the often high spatial heterogeneity of surface fires. Fuel moisture dynamics of leaf litter are dependent upon forest structure (shading and rain throughfall effects), litter physical structure, and chemical composition of the fuelbeds \cite{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}.

Fuel moisture content through time is controlled by two main processes, absorptive capacity and rate of moisture loss \citep{Kreye-2013}. Absorptive capacity represents the potential maximum fuel moisture and is evaluated by the moisture content of litter when saturated. After saturation, the differences in moisture content indicate different abilities to retain moisture. The ability of litter to absorb and desorb water should be dependent on leaf and/or litter geometry, however, which traits control these processes are largely unknown. We propose two different sets of traits responsible for variation in the two processes mentioned: leaf particle geometry and litter bulk density. Litter characteristics such as surface-area-to-volume ratio, litter bulk density, mass density, particle size and shape, and chemical traits influence moisture diffusion through litter. Litter packing has two pathways of effect on fire behavior: directly by its effect on oxygen availability for combustion, and indirectly by its effect on dry-down rate, and both these pathways are affected by leaf litter traits (such as leaf size, litter bulk density). Fuelbeds composed of larger leaves with lower moisture retainability and decomposition rates favor drying, thus increasing the probability of ignition and fire spread rate.

Species traits may interact such that the flammability of multi-species mixtures is not the average of the constituent species.  Such non-additive effects have been shown to be synergistic in dry litter fuels where the flammability of a mixture is driven by the most flammable species  \cite{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}. In most forests, leaf litter will be a mixture representative of the species in the stand. It is possible that similar non-additive effects may influence litter moisture content, especially if different leaf traits influence adsorptive capacity and drying rate. Blauw et al.\cite{Blauw+Wensink+etal-2015} suggests that moisture content can contribute to non-additive effects on heathland flammability, but the mechanism of this effect is not known.

% I investigated non-additive effects on both moisture content and the flammability of litter mixtures varying in time since wetting and in moisture content. I expect that litter mixtures will differ in their bulk density and thus, will dry at different rates.

We investigated species trait effects on litter moisture dynamics and consequent litter flammability across eight tree species in a temperate fire-prone forest in California. We determined the rate of moisture loss, the moisture of extinction (the threshold representing the moisture value above which fire will not ignite or spread) and the influence of moisture on the flammability of eight species from a mixed-conifer forest in California. Experimental burn trials of leaf litter at various moisture levels determined how time since wetting influenced time to ignition and flame spread rate. The burn trials were also performed with multi-species litter mixtures to evaluate if species contribute equally to litter moisture dynamics or if there are non-additive effects. We compared leaf litter traits with the measured moisture absorption and desorption rates. Finally, we tested if the non-additive effects on flammability reported in previous studies occur under ecologically-relevant moisture levels and whether fuel moisture effects strengthen or mask such non-additivity.

@@

Field sites for this study were located in Sequoia and Kings Canyon National Parks, California, USA (36 36N, 118 42 W) between 1600 and 2400 m elevation in mixed-conifer forest. I chose eight tree species that are representative of this type of forest: \emph{Pinus jeffreyi} Grev. \& Balf., \emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws., \emph{Abies concolor} (Gord. \& Glend.) Lindl. ex Hildebr., \emph{Abies magnifica} A. Murr., \emph{Calocedrus decurrens} (Torr.) Florin, \emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J. Buchholz. In this study, I consider litter the top layer of leaves and small twigs less than 0.625 cm diameter (= 1-hour fuel) that have fallen in the previous year (mostly undecomposed). I collected the leaf litter in the summer (mid-June to mid-July) in 2011 and in 2012 from 21 separate sites across the study area. Effort was made to collect from different populations (minimum of four, each population separated from the others by 9 to 32 km) for each species to control for plastic and ecotypic effects. An exception to this involved the litter from A. magnifica which is restricted to the higher elevations and could only be collected from one area, due to logistical constraints. Within that area, \emph{A. magnifica} litter was collected from over 10 different individuals. For the remaining species, the collection involved 2-4 individual trees at each site, separated by at least 10 m. The litter was collect about 2 m away from the trunk to obtain a more uniform sample, because bark and heavier branches tend to fall closer to the tree trunk.

@@

We produced dry-down curves to determine maximum water retainability and the rate of moisture content loss (dessication rate) for each species and for four distinct mixtures. These curves allowed us to  determine the rate at which a species loses moisture (approximated as
exponential decay), and determine maximum water content following saturation
and gravity draining. Samples from the eight species (six replicates per
species) and four mixtures (5 replicates per mixture type) were placed in
baskets (approximately 45 x 45 cm by 15 cm tall) constructed of 1 cm aluminum
screen. Samples were 450 g of litter (dry weight) per trial, and litter depth
was standardized to 10 cm by changing the horizontal dimensions of the basket.
Weight was measured by using a balance sensitive to 0.1 g (model XS16001L,
Mettler Toledo, Columbus, OH). Samples were immersed in plastic storage bins
filled with water for 24 hours to fully saturate them. After saturation, the
baskets were removed from the water, allowed to drain for 3 minutes, and
samples weighed to measure maximum moisture content. The samples were then
allowed to dry in a partially controlled environment kept at 21 C temperature
and 30\% relative humidity, assessed via an iButton sensor (Maxim Integrated)
placed over the samples. During the dry-down process three subsamples were
taken at 24 hour intervals and weighed. These subsamples were then oven dried
for 24 hours at 100 C and re-weighed to assess moisture content. This was
repeated until the final sample from a basket was less than10\% moisture.

The dry-down curves indicated that the eight species varied in their moisture behavior [FIG 2?]. To determine a subset of species to use in the mixture trials, we performed analysis of contrasts using lsmeans package in R  (TODO Lenth 2016) to determine similarity between the species, which fell into four groups. I then selected one species from each group to include in the mixtures resulting in four distinct 3-species mixtures.

 Leaf trait data

We investigated the influence of leaf thinness and litter bulk density on the moisture dynamics of the species in this study. Leaf thinness data (leaf length over thickness) was obtained for subsamples of the litter collected for this project.  Bulk density of the litter was obtained for each basket during the dry-down experiment by dividing the dry mass (450 g) by the estimated litter volume (basket width x length x litter depth).

@@

We performed flammability tests across multiple moisture levels on monospecific litter beds and on mixed litter beds with eaach species or mixture replicated 5 times at each moisture level. Flammability was assessed using a 150 cm long burn table, 15 cm wide and 15 cm tall, in which leaf litter was placed and the table gently shaken to allow settling. This design allows fire behaviour to reach constant flame spread rates \cite{Magalhaes+Schwilk-2012}. Two graduated metal rulers   equally spaced along the apparatus allowed visual determination of maximum flame height.
Litter was ignited with a propane torch; maximum adiabatic flame temperature of 1899 C, http://www.benzomatic.com/products/fuel.aspx) at one end of the apparatus and allowed to burn until extinction. If a species at a specific moisture level failed to ignite after three attempts, no further attempts were made.  A timer was used to record time to ignition and duration of flaming combustion. The flammability trials followed the methodology detailed in de Magalhães and Schwilk \cite{Magalhaes+Schwilk-2012}.

% and nine type-K thermocouples attached to data loggers measured temperature every second at three distances from the ignition point and at three different depths below the litter surface.

During flammability trials, we recorded time to ignition (s), duration of flaming combustion (s), spread rate (mm/s), calculated by dividing the length of the burned surface by the time it takes the flaming front to reach the end of the apparatus, maximal flame height (mm), and mass loss (\%), calculated as the difference between fuel mass before and after flame extinction. I also measured duration above 100 C, or the length of time temperature exceeded 100 C, and temperature integration, which integrates duration of combustion and temperature above 100 C and serves as a proxy for heat release (de Magalhães and Schwilk 2012, Cornwell et al. 2015). By performing the flammability tests at several moisture levels I were able to determine moisture of extinction (a threshold representing the moisture value above which litter will not ignite or spread) for all species in this study, using a logistic regression.

Burn trials occurred in a cement structure used to simulate house fires at the Fire Department of the City of Lubbock.This structure minimizes wind and helps regulate temperature and relative humidity. Temperature and relative humidity were measured before every trial using a Kestrel 3000 (Nielsen-Kellerman, Boothwyn, PA). To minimize temperature and relative humidity variation, all trials were conducted on clear days between 10 am and 3 pm.

@@

Dry-down curves were created based on the drying experiment, fitted with an exponential decay curve to each species.:

\begin{equation}
m = m_{max} e^{-\lambda t}
\end{equation}

Where $m$ is moisture content at time $t$, and $\lambda$ is the exponential decay coefficient. To estimate these c oefficents we fit a linear mixed-effects model by first taking the natural logarithm of moisture content, with time since wetting and species as fixed effects nested within replicate basket. We extracted the coefficients of the fitted model to describe maximum water retainability ($m_{max}$; g of water per g of dry mass) and desiccation rate ($\lambda$) for each species. All analyses were conducted in R \cite{RCoreTeam-2019}. We fit models with R packages `lme4' \cite{Bates_Machler_etal-2015} and `afex' \cite{Singmann_Bolker_etal-2017}.  Approximate degrees of freedom and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997} which is recommended by Luke \cite{Luke-2017} as producing acceptable Type 1 error rates.

We tested if litter bulk density influenced moisture dynamics (maximum water retainability and desiccation rate) by fitting linear models with species means as observations.

The effect of time since wetting on flammability was modelled using linear regression with the interaction between time since wetting and genus as fixed effects and average vapor pressure deficit in the burning room during that trial as a nuisance covariate. Differences across species were tested with post-hoc Tukey tests. A binomial logistic regression was fitted to the probability of ignition to obtain moisture of extinction.

We investigated non-additive effects on moisture and on flammability by testing if moisture content and flammability of mixtures were predicted by the average of the constituent species. To choose mixtures we first grouped the species according to their maximum water retainability and dessication rate according to estimated marginal means by species using the R package `emmeans' \cite{Lenth-2019}. This resulted in four broad groups of species and we then selected one representative species from each litter type group and produced all four possible three-species litter mixtures using those four species. For each mixture and time since wetting, we predicted moisture content and flame spread rate using the models fit to single species data described above and then calculated an expected moisture content and spread rate as the average of these three values for the mixture. This procedure was done for both moisture content and flammability parameter values of the constituent species. The expectation from the null model was that the difference between observed and predicted values of flammability will be zero. We tested if there were significant departures from the null (zero) with a mixed effects linear model.

@@

There are marked differences in the extent of moisture species retain initially following wetting and draining by gravity (all p < 0.0001, Fig. \ref{fig1-drydown} and Table \ref{tab1-drydown}). \emph{Q. kelloggii}, the only broadleaf, is the species that retains the most moisture but has had a rapid dessication rate (. Calculating estimated marginal means following the linear mixed-effects model discriminated the species into three different maximum water retention groups: 1) \emph{Q. kelloggii}; 2) \emph{P. lambertiana} and \emph{C. decurrens}; and 3) \emph{A. magnifica}, \emph{S. giganteum}, \emph{A. concolor}, \emph{P. jeffreyi} and \emph{P. ponderosa}.  There are also differences in desiccation rates across the eight species and estimated mixed linear model slopes discriminated species into three groups from lowest to most rapid: 1) (\emph{C.decurrens}, 2) \emph{Abies spp} and \emph{S. gianteum}, and 3) \emph{Q. kelloggii} and \emph{Pinus spp}.

\begin{figure}[H]
  \centering
  \label{fig-drydown}
\includegraphics[width=8cm]{figs_tables/fig1_drydown-curves.pdf}
\caption{Dry down curves for eight litter types. Lines show best-fit exponential decay curves fit to moisture content on a dry mass basis. Litter samples we saturated by immersion for 24 hours and then allowed to drain by gravity for 3 min before initial weighing. Colors indicate genera and there were significant differences in both }
\end{figure}

\begin{table}[H]
  \caption{Mixed model results for moisture content as a function of time since wetting. Approximate degrees of freedom, pseudo F statistics and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.}
  \label{tab1-drydown}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\input{figs_tables/drydown-tab.ltx}
\end{table}

Across these eight species litter bulk density had a negative effect on desiccation rate (Fig. \ref{fig-bd-di}, linear model p = 0.011).

\begin{figure}[H]
  \centering
  \label{fig-bd-di}
\includegraphics[width=8cm]{figs_tables/fig3_di_by_bd.pdf}
\caption{ }
\end{figure}

@@

Time to ignition and spread rate were modeled against time since wetting and using genus as a fixed effect.

[TODO]?  redo in terms of groups.

@@

Based on visual inspection of the dry down curves and the estimated marginal means for maximum water retainability and desiccation rate, selected four species to represent the full range of litter moisture dynamics for use in the litter mixtures: \emph{Abies concolor}, \

@@
